Chemistry of the Orphan Stream I: Identifying Stream Members from Low Resolution Spectroscopy

### 2012-09-15 2012-oss-mrs/draft-oss-mrs.tex
Fixed small latex compilation bug
Added stream metallicity distribution text

In this letter we present a detailed, self-consistent analysis to identify K giant members of the Orphan Stream. Using our selection method we have catalogued the locations of X highly probable Orphan Stream candidates, all worthy of high resolution spectroscopic follow up. In the following section we outline our photometric target selection. In \S\ref{sec:observations} we describe the low resolution spectroscopic observations. The data analysis, including stream identification, is discussed in \S\ref{sec:analysis} and in \S\ref{sec:conclusions} the conclusions, predictions and future work are presented.

@@

We have measured the equivalent widths of the gravity-sensitive Mg I line at 8807 \AA{} to distinguish dwarfs from giants \citep{Battaglia;Starkenburg_2012}. At a given temperature (or $g \-- r$) and metallicity, giant stars present narrower Mg I absorption lines than their dwarf contaminants. Given our target selection, our sample is likely to contain many more dwarfs than giants. In some cases no Mg I 8807 \AA{} line was apparent, so an upper limit was estimated based on the S/N. In these cases the candidate was considered a ``non-dwarf" because we cannot exclusively rule out a metal-poor sub-giant with this criterion alone. For these purposes we are only looking for a simple indication as to whether a star is likely a dwarf or not. Figure \ref{fig:ew-mg} illustrates the trend with EW$_{\lambda8807}$ against SDSS de-reddened $g \-- r$\footnote{All magnitudes presented in this letter are de-reddened}, illustrating the dominant upper dwarf branch we wish to exclude. The adopted dwarf/giant separation line is shown. Admittedly this selection criterion is not particularly strict, but we are allowing multiple inclusive cuts and taking the intersection of these criteria to identify Orphan Stream giants.

@@

We have measured the equivalent widths of the Ca\,\textsc{II} near-infrared triplet lines for stars that meet our kinematic and surface gravity criteria. After correcting for luminosity, the equivalent width of the Ca\,\textsc{II} triplet lines provide a good indication of the metallicity of a RGB star \citep{Amandroff;Da_Costa_1991}. We have employed the \citet{Starkenburg;et-al_2010} relationship and corrected for luminosity in $g$ against the horizontal branch magnitude at $g_{HB}$ = 17.1 \citep{Newberg;et-al_2010}. Strictly speaking, the Ca\,\textsc{II}\---[Fe/H] calibration is only valid for stars brighter than the horizontal branch, although the relationship only becomes significantly inappropriate near $g \-- g_{HB} \sim +1$ \citep{Saviane;et-al_2012}. Many of our candidates are fainter than this valid luminosity range, and therefore they should not be excluded solely because of their derived metallicities. Stars fainted than $g_{HB}$ will have slightly lower metallicities than predicted by our Ca\,\textsc{II}\---[Fe/H] relationship, and for these stars we will only use metallicities to assign a relative qualitative likelihood for stream membership.

Given a distance to the Orphan Stream, we can also estimate a star's metallicity through isochrone fitting. We have used a 10 Gyr \citet{Girardi;et-al_2008} isochrone at 21.4\,kpc \citep{Newberg;et-al_2010} and found metallicities for all eighteen likely stream members from their best-fitting isochrone (see Figure \ref{fig:feh}). Derived metallicites from Ca\,\textsc{II} line strengths and isochrone fitting which are consistent (within $\pm0.3$\,dex) indiates these measurements are reliable, and that these stars are indeed at a distance of $\sim21.4$\,kpc. These ten highly likely stream members fall in the green shaded region illustrated in Figure \ref{fig:feh}.

TODO: Overall stream metallicity

\begin{figure}[h]
	\includegraphics[width=\columnwidth]{./figures/feh.pdf}
	\caption{Metallicities from the Ca \textsc{II} triplet lines against metallicities derived from isochrone fitting for the eighteen stars which meet our kinematic and surface gravity criteria. Both abundance determinations imply these stars are RGB members of the Orphan Stream at a distance of $\sim21.4$ kpc \citep{Newberg;et-al_2010}. Consistency between these methods indicates highly likely stream membership (green region). The minimum isochrone metallicity, and a representative uncertainty of 0.2\,dex for abundance measurements is shown.}
	\label{fig:feh}
\end{figure}

Figure \ref{fig:cmd} illustrates all eighteen candidates which meet both our kinematic and surface gravity criteria, and our ten most likely members are marked. Candidates are coloured by their metallicity deduced from the Ca\,\textsc{II} triplet.

\begin{figure}[h]
	\includegraphics[width=\columnwidth]{./figures/cmd.pdf}
	\caption{Color magnitude diagram showing our observed candidates (grey). Observations fulfilling kinematic and gravity cuts are colored by their metallicity. Those with upper limits for surface gravity are marked as triangles. Highly probable stream members with consistent line strength- and isochrone-derived abundances are circled. Relevant 10 Gyr \citet{Girardi;et-al_2008} isochrones at $[\mbox{Fe/H}] = -1.5$ (dotted), $-2.0$ (dashed) at 21.4\,kpc \citep{Newberg;et-al_2010} are shown, as well as our best-fitting 10 Gyr isochrone of $[\mbox{Fe/H}] = -1.78$ at 24\,kpc (solid).}
	\label{fig:cmd}
\end{figure}

\citet{Newberg;et-al_2010} found the median Orphan Stream metallicity to be [Fe/H] $= -2.1$ from BHB stars, and a distance of $21.4 \pm 1.0$\,kpc in this region. In Figure \ref{fig:cmd} we have overlaid 10 Gyr old isochrones from \citet{Girardi;et-al_2008} with metallicities of [Fe/H] $= -2.0$ and $-1.5$ at a distance of $21.4$ kpc \citep{Newberg;et-al_2010}.

%We can ascertain abundance corrections for our stars fainter than the horizontal branch by comparing our $g - g_{HB}$ (or $V-V_{HB}$) and $EW_{\lambda8452} + EW_{\lambda8662}$ with those measured by \citet{Saviane;et-al_2012} for globular clusters. Only two clusters from \citet{Saviane;et-al_2012} had measured Ca \textsc{II} equivalent widths for stars fainter than the horizontal branch: NGC 6397 with $[\mbox{Fe/H}] = -1.99$ and NGC 6254 with $[\mbox{Fe/H}] = -1.57$ \citep[both abundances from][]{Carretta;et-al_2009}. OSS 6 and OSS 7 lie at the mid-point between these cluster stars, implying an abundance of $[\mbox{Fe/H}] \sim -1.78$, or a mean correction of $-0.14$ dex for OSS 6 and 7. Although OSS 3 is 0.15 magnitudes fainter than the horizontal branch, it has equivalent widths consistent with our inferred metallicity of $[\mbox{Fe/H}] = -1.62$, so no correction was made. With corrections included for OSS 6 and 7 we find a median stream metallicity from our four most probable stars of $[\mbox{Fe/H}] = -1.78 \pm 0.08$, which is slightly more metal-rich than the $[\mbox{Fe/H}] =-2.1$ found by \citet{Newberg;et-al_2010} from BHB stars. If we include our EMP candidate, the median decreases by 0.01 dex and the spread increases to 0.39\,dex.

Given that our Orphan Stream stars cover a wide range along the giant branch, we are in a good position to revise the distance estimate to the stream. Given a 10 Gyr \citet{Girardi;et-al_2008} isochrone at $[\mbox{Fe/H}] = -1.78$, we find a best-fitting distance to the stream of 24 kpc at ($l, $b) = ($250^\circ$, $50^\circ$). This isochrone is also shown in Figure \ref{fig:cmd}, and it is clear that the four most probable member stars lie almost precisely on this isochrone. Our derived distance is slightly farther than the $21.4 \pm 1.0$ kpc found by \citet{Newberg;et-al_2010}, but given the inherent uncertainties in determining distances we argue these measurements are largely in agreement.

\begin{deluxetable*}{lccccccccccc}
\tabletypesize{\scriptsize}
\tablecaption{Identified Orphan Stream Members\label{tab:oss-members}}
\tablehead{
	\colhead{Star} &
	\colhead{$\alpha$} &
	\colhead{$\delta$} &
	\colhead{$g$} &
	\colhead{$g \-- r$} &
	\colhead{$V_{helio}$} &
	\colhead{$V_{\textsc{gsr}}$} &
	\colhead{$V_{err}$} &
	\colhead{$EW_{\lambda8807}$} &
	\colhead{[Fe/H]$_{Ca}$} &
	\colhead{[Fe/H]$_{iso}$} &
	\colhead{Stream} \\
Name & (J2000) & (J2000) & & & (kms $^{-1}$) & (km s$^{-1}$) & (km s$^{-1}$) & (\AA) & (dex) & (dex) & Probability
}
\startdata
OSS 1  & 10 46 29.3 & $-$00 19 38.5 & 16.97 & 0.56 & ? &  78.4 &  5.2 &    0.126 & --1.63 &    --1.68 & High \\
OSS 2  & 10 47 06.1 & $-$01 56 03.9 & 17.98 & 0.54 & ? &  74.9 & 17.6 &    0.452 & --1.12 &    --1.40 & High \\
OSS 3  & 10 47 17.6 & $+$00 25 07.7 & 14.89 & 0.73 & ? &  79.2 &  3.0 &    0.213 & --1.77 &    --1.80 & High \\
OSS 4  & 10 47 29.1 & $-$02 02 22.6 & 17.20 & 0.46 & ? &  93.2 & 29.8 & $<$0.22 & --2.82 & $<$--2.28 & High \\
OSS 5  & 10 47 30.1 & $-$00 01 24.6 & 17.25 & 0.61 & ? &  83.5 &  3.2 &    0.123 & --1.62 &    --1.68 & High \\
OSS 6  & 10 48 20.9 & $+$00 26 34.4 & 17.88 & 0.55 & ? & 118.8 & 11.7 &    0.467 & --1.65 &    --1.73 & High \\
OSS 7  & 10 48 44.4 & $-$02 53 08.8 & 18.35 & 0.62 & ? & 108.1 &  9.0 &    0.183 & --1.01 &    --1.17 & High \\
OSS 8  & 10 48 46.9 & $-$00 32 27.8 & 17.85 & 0.46 & ? & 109.3 &  8.1 &    0.324 & --2.37 & $<$--2.28 & High \\
OSS 9  & 10 49 08.3 & $+$00 02 00.2 & 16.27 & 0.62 & ? &  81.5 &  4.6 &    0.034 & --2.70 & $<$--2.28 & High \\
OSS 10 & 10 51 19.7 & $+$00 05 15.4 & 17.81 & 0.65 & ? &  82.6 &  5.0 &    0.198 & --1.16 &    --1.20 & High 
\enddat
\end{deluxetable*}

%OSS 1 & 10 46 50.6 & $-$00 13 17.9 & 17.33 & 0.50  &  215.5  & 77.0 & 4.0 & 0.416 & $-1.31$ & Unlikely member! \\
%OSS 2 & 10 47 17.6 & $+$00 25 07.7 & 16.09 &  0.73 &  215.7  & 79.2 & 3.3 & 0.212 & $-1.84$ & Highly probable member \\
%OSS 3 & 10 47 30.1 & $-$00 01 24.6 & 17.25 &  0.61 &  221.4  & 83.6 & 3.4 & 0.123 & $-1.62$ & Highly probable member \\
%OSS 4 & 10 49 08.3 & $+$00 02 00.2 & 16.27 &  0.62 &  218.5  & 81.5 & 4.6 & 0.034 & $-2.70$ & Possible EMP member \\
%OSS 6 & 10 48 20.9 & $+$00 26 34.4 & 17.88 &  0.55 &  254.9  & 118.8 & 11.7 & 0.469 & $-1.79$\tablenotemark{a} & Highly probable member \\
%OSS 7 & 10 46 29.3 & $-$00 19 38.5 & 17.77 &  0.56 &  217.3  & 78.4 & 5.2 & 0.126 & $-1.77$\tablenotemark{a} & Highly probable member

@@

We have presented a detailed analysis to isolate individual Orphan Stream K giants from low resolution spectroscopy using a combination of photometric, kinematic, gravity, and metallicity information. Although each individual criterion is likely to induce some level contamination, their intersection reveals highly probable, self-consistent, Orphan Stream K giants.  We deduce a stream metallicity of $[\mbox{Fe/H}] = -1.78 \pm 0.08$ which is not biased by spectral type. Recall that the metallicity determination was performed after kinematic and gravity cuts, and our most probable members lay perfectly on a 10 Gyr isochrone of $[\mbox{Fe/H}] = -1.78$. Our stream members indicate a revised distance to the stream of $24$\,kpc at $(l, b) = (250\,^\circ$, $50\,^\circ)$, slightly farther than that deduced by \citet{Newberg;et-al_2010}.

@@

Using a maximum-likelihood estimation we find the mean stream galactocentric velocity at $(l, b) = (250\,^\circ, 50\,^\circ)$ from five stars to be $V_{GSR} = 82 \pm 1.6$ km s$^{-1}$ and the intrinsic dispersion to be $0.3 \pm 3.3$ km s$^{-1}$, i.e., consistent with zero. If we exclude five stars with weak signal \--- and hence large velocity uncertainties \--- this dispersion is found to be $X \pm Y$\,km s$^{-1}$.

For our four stars with low radial velocity errors this dispersion is found to be $0.00 \pm 2.4$ km s$^{-1}$. These velocities indicate that the Orphan Stream is a very cold stream, and implies their motion is largely perpendicular to the line of sight.

>>>>

In this letter we present a detailed, self-consistent analysis to identify K giant members of the Orphan Stream. Using our selection method we have catalogued the locations of ten highly probable Orphan Stream candidates, all worthy of high resolution spectroscopic follow up. In the following section we outline our photometric target selection. In \S\ref{sec:observations} we describe the low resolution spectroscopic observations. The data analysis, including stream identification, is discussed in \S\ref{sec:analysis} and in \S\ref{sec:conclusions} the conclusions, predictions and future work are presented.

@@

We have measured the equivalent widths of the gravity-sensitive Mg I line at 8807 \AA{} to distinguish dwarfs from giants \citep{Battaglia;Starkenburg_2012}. At a given temperature (or $g \-- r$) and metallicity, giant stars present narrower Mg I absorption lines than their dwarf contaminants. Given our target selection, our sample is likely to contain many more dwarfs than giants. In some cases no Mg I 8807 \AA{} line was apparent, so an upper limit was estimated based on the S/N. In these cases the candidate was considered a ``non-dwarf" because we cannot exclusively rule out a metal-poor sub-giant with this criterion alone. For these purposes we are only looking for a simple indication as to whether a star is likely a dwarf or not. Figure \ref{fig:ew-mg} illustrates the trend with EW$_{\lambda8807}$ against SDSS de-reddened\footnote{All magnitudes presented in this letter are de-reddened} $g \-- r$, illustrating the dominant upper dwarf branch we wish to exclude. The adopted dwarf/giant separation line is shown. Admittedly this selection criterion is not particularly strict, but we are allowing multiple inclusive cuts and taking the intersection of these criteria to identify Orphan Stream giants.

@@

We have measured the equivalent widths of the Ca\,\textsc{II} near-infrared triplet lines for stars that meet our kinematic and surface gravity criteria. After correcting for luminosity, the equivalent width of the Ca\,\textsc{II} triplet lines provide a good indication of the overall metallicity of a RGB star \citep{Amandroff;Da_Costa_1991}. We have employed the \citet{Starkenburg;et-al_2010} relationship and corrected for luminosity in $g$ against the horizontal branch magnitude at $g_{HB}$ = 17.1 \citep{Newberg;et-al_2010}. Strictly speaking, the Ca\,\textsc{II}\---[Fe/H] calibration is only valid for stars brighter than the horizontal branch, although the relationship only becomes significantly inappropriate near $g \-- g_{HB} \sim +1$ \citep{Saviane;et-al_2012}. Many of our candidates are fainter than this valid luminosity range, and therefore they should not be excluded solely because of their derived metallicities. Stars fainted than $g_{HB}$ will have slightly lower metallicities than predicted by our Ca\,\textsc{II}\---[Fe/H] relationship, and for these stars we will only use metallicities to assign a relative qualitative likelihood for stream membership.

\begin{figure}[h]
	\includegraphics[width=\columnwidth]{./figures/feh.pdf}
	\caption{Metallicities from the Ca \textsc{II} triplet lines against metallicities derived from isochrone fitting for the eighteen stars which meet our kinematic and surface gravity criteria. Both abundance determinations imply these stars are RGB members of the Orphan Stream at a distance of $\sim21.4$\,kpc \citep{Newberg;et-al_2010}. Consistency between these methods indicates highly likely stream membership (green region). The minimum isochrone metallicity, and a representative uncertainty of 0.2\,dex for abundance measurements is shown.}
	\label{fig:feh}
\end{figure}

Given a distance to the Orphan Stream, we can also estimate a star's metallicity through isochrone fitting. We have used a 10 Gyr \citet{Girardi;et-al_2008} isochrone at 21.4\,kpc \citep{Newberg;et-al_2010} and found metallicities for all eighteen likely stream members from their best-fitting isochrone (see Figure \ref{fig:feh}). Derived metallicities from Ca\,\textsc{II} line strengths and isochrone fitting which are consistent (within $\pm0.3$\,dex) indicates these measurements are reliable, and that these stars are indeed at a distance of $\sim21.4$\,kpc. We find ten highly likely stream members with consistently derived metallicities, and they fall within the green shaded region illustrated in Figure \ref{fig:feh}.

We have adopted a final overall metallicity for each star based on the quality of our [Fe/H] measurements. Adopted values are tabulated in Table \ref{tab:oss-members}. From our ten highly likely stream members we find an overall stream metallicity of $[\mbox{Fe/H}] = -1.71$ with a dispersion of $\sigma = 0.56$ dex. This abundance spread is larger than typically seen in globular cluster stars and is more representative of the chemical spreads seen in dSph satellites. \citet{Newberg;et-al_2010} found a stream metallicity of $[\mbox{Fe/H}] = -2.1$ from BHB stars, which is not inconsistent with our measurement given the spread in Orphan Stream K-giant metallicities.

\begin{figure}[h]
	\includegraphics[width=\columnwidth]{./figures/cmd.pdf}
	\caption{Color magnitude diagram showing our observed candidates (grey). Observations fulfilling kinematic and gravity cuts are colored by their metallicity, and those with upper limits for surface gravity are marked as triangles ($\triangledown$). Highly probable stream members with consistent line strength- and isochrone-derived metallicities are circled. Relevant 10 Gyr \citet{Girardi;et-al_2008} isochrones at $[\mbox{Fe/H}] = -1.5$ (dotted), $-2.0$ (dashed) at 21.4\,kpc \citep{Newberg;et-al_2010} are shown, as well as our best-fitting 10 Gyr isochrone of $[\mbox{Fe/H}] = -1.71$ at YY\,kpc (solid).}
	\label{fig:cmd}
\end{figure}

Given that our Orphan Stream giants cover a wide evolutionary range along the giant branch (Figure \ref{fig:cmd}), we are in a good position to revise the distance estimate to the stream. Given a 10 Gyr \citet{Girardi;et-al_2008} isochrone at $[\mbox{Fe/H}] = -1.71$, we find a best-fitting distance to the stream of YY\,kpc at ($l, $b) = ($250^\circ$, $50^\circ$). This isochrone is also shown in Figure \ref{fig:cmd}, and it is clear that the four most probable member stars lie almost precisely on this isochrone. Our derived distance is slightly farther than the $21.4 \pm 1.0$ kpc found by \citet{Newberg;et-al_2010}, but given the inherent uncertainties in determining distances we argue these measurements are largely in agreement.

\begin{deluxetable*}{lcccccccccccc}
\tabletypesize{\scriptsize}
\tablecaption{Identified Orphan Stream Members\label{tab:oss-members}}
\tablehead{
	\colhead{$\alpha$} &
	\colhead{$\delta$} &
	\colhead{$g$} &
	\colhead{$g \-- r$} &
	\colhead{$V_{hel}$} &
	\colhead{$V_{\textsc{gsr}}$} &
	\colhead{$V_{err}$} &
	\colhead{$EW_{\lambda8807}$} &
	\colhead{[Fe/H]$_{Ca}$} &
	\colhead{[Fe/H]$_{iso}$} &
	\colhead{[Fe/H]\tablenotemark{a}} &
	\colhead{Stream} \\
(J2000) & (J2000) & & & (kms $^{-1}$) & (km s$^{-1}$) & (km s$^{-1}$) & (\AA) & (dex) & (dex) & (dex) & Prob.
}
\startdata
10 46 29.3 & $-$00 19 38.5 & 16.97 & 0.56 & ? &  78.4 &  5.2 &    0.126 & --1.63  &    --1.68 & --1.63 & High \\
10 47 06.1 & $-$01 56 03.9 & 17.98 & 0.54 & ? &  74.9 & 17.6 &    0.452 & :--1.12\tablenotemark{b} &    --1.40 & --1.40 & High \\
10 47 17.6 & $+$00 25 07.7 & 14.89 & 0.73 & ? &  79.2 &  3.0 &    0.213 & --1.77  &    --1.80 & --1.77 & High \\
10 47 29.1 & $-$02 02 22.6 & 17.20 & 0.46 & ? &  93.2 & 29.8 & $<$0.22  & --2.82  & $<$--2.28 & --2.82 & High \\
10 47 30.1 & $-$00 01 24.6 & 17.25 & 0.61 & ? &  83.5 &  3.2 &    0.123 & --1.62  &    --1.68 & --1.68 & High \\
10 48 20.9 & $+$00 26 34.4 & 17.88 & 0.55 & ? & 118.8 & 11.7 &    0.467 & --1.65  &    --1.73 & --1.73 & High \\
10 48 44.4 & $-$02 53 08.8 & 18.35 & 0.62 & ? & 108.1 &  9.0 &    0.183 & :--1.01\tablenotemark{b} &    --1.17 & --1.17 & High \\
10 48 46.9 & $-$00 32 27.8 & 17.85 & 0.46 & ? & 109.3 &  8.1 &    0.324 & --2.37  & $<$--2.28 & --2.37 & High \\
10 49 08.3 & $+$00 02 00.2 & 16.27 & 0.62 & ? &  81.5 &  4.6 &    0.034 & --2.70  & $<$--2.28 & --2.70 & High \\
10 51 19.7 & $+$00 05 15.4 & 17.81 & 0.65 & ? &  82.6 &  5.0 &    0.198 & --1.16  &    --1.20 & --1.20 & High 
\enddata
\tablenotetext{a}{Final adopted [Fe/H] value based on quality of our derived metallicities.}
\tablenotetext{b}{Sufficiently fainter than $g_{HB}$ to qualify this measurement as uncertain.}
\end{deluxetable*}

@@

We have presented a detailed analysis to isolate individual Orphan Stream K giants from low resolution spectroscopy using a combination of photometric, kinematic, gravity, and metallicity information. Although each individual criterion is likely to induce some level contamination, their intersection reveals highly probable, self-consistent, Orphan Stream K giants.  We deduce a median stream metallicity of $[\mbox{Fe/H}] = -1.71$ which is not biased by spectral type, and find an intrinsically wide metallicity spread of $\sigma = 0.56$\,dex, indicative of a dSph origin. Recall that the metallicity determination was performed after kinematic and gravity cuts, and our most probable members lay perfectly on a 10 Gyr isochrone of $[\mbox{Fe/H}] = -1.71$. Our stream members indicate a revised distance to the stream of $YY$\,kpc at $(l, b) = (250\,^\circ$, $50\,^\circ)$, Z\,kpc slightly farther than that deduced by \citet{Newberg;et-al_2010}.

@@

Using a maximum-likelihood estimation we find the mean stream galactocentric velocity at $(l, b) = (250\,^\circ, 50\,^\circ)$ from five stars to be $V_{GSR} = ZZ.Z \pm Z.Z$ km s$^{-1}$ and the intrinsic dispersion to be $Z.Z \pm Z.Z$ km s$^{-1}$, i.e., consistent with zero. If we exclude five stars with weak signal \--- and hence large velocity uncertainties \--- this dispersion is found to be $Z \pm Z$\,km s$^{-1}$. These velocities indicate that the Orphan Stream is a very cold stream, and implies their motion is largely perpendicular to the line of sight.
